 

### 2023-02-23 paper3.Rmd
Edits to Paper 3

    text = element_text(family = "Source Sans Pro")) +
  scale_color_manual(values = c(col_mh,
                                col_headache,
                                col_grey_lightest),
                     name = "") +
  labs(x = "Slope change", y = "Level shift") +
  geom_text_repel(data=subset(model_outputs_its_graph, estimate_slope < -0.04 |
                                (estimate_level <.1 & mh_indicator ==  "Mental symptom" |
                                   mh_indicator == "Headache (control)")), max.overlaps = 10,
                  size = 3)

@@

While our ITS analysis provided suggestive evidence that search density for mental health symptoms might have changed in the post-announcement period, it also highlighted that searches may have generally changed across other diseases at the same time. Our CITS estimates of each mental health symptom (Figure \ref{figure_cits_outputs}) using searches for headaches as a comparison broadly echoed our ITS analysis, with most conditions showing level and slope changes close to the null. For three conditions, our CITS analysis showed relatively large slope changes with searches for anxiety `r function_dynamic_text_results("main","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("main","depression", "slope", value_include = "allnumber")`, and MDD `r function_dynamic_text_results("main","major_depressive_disorder", "slope", value_include = "allnumber")` all showing sustained declines. All three of these symptoms had relatively large slope changes in our ITS analysis. In addition, the CITS level shift (i.e., the relative immediate effect) for anxiety was `r function_dynamic_text_results("main","anxiety", "level")`, a decrease double in magnitude compared to the majority of estimates for other symptoms.

@@

```{r figures_cits_data_anxiety, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 5.5, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for anxiety, by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_anxiety}Density of Google Search trends for anxiety, by region and as a function of time. In each row, search density data for two symptoms are displayed: anxiety in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```
```{r figures_cits_data_depression, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 5.5, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for depression, by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_depression}Density of Google Search trends for depression, by region and as a function of time. In each row, search density data for two symptoms are displayed: depression in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r figures_cits_data_mdd, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 5.5, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for major depressive disorder (MDD), by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_mdd}Density of Google Search trends for MDD, by region and as a function of time. In each row, search density data for two symptoms are displayed: MDD in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

Examining the plausible model combinations of the multiverse analysis for these three symptoms underscored the importance of regional level to inference, as the use of county-level data (with a greater number of observations) consistently showed similar effect estimates to state level estimates for each CITS estimand, but with much tighter confidence intervals, even when geographic fixed effects and narrower bandwidths were included. Slope changes and level shifts for anxiety, depression, and MDD are reported for each combination (144 models) in Figure \ref{fig_multiverse_analysis}. Across all models, the sustained change in search density for anxiety and depression is negative and significant an alpha of 0.01. In terms of the estimated level shifts, `r function_dynamic_text_results("multiverse","anxiety", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)` models for anxiety (`r round(function_dynamic_text_results("multiverse","anxiety", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)*100/144,0)`%) had a statistically significant negative change, compared to `r function_dynamic_text_results("multiverse","anxiety", "intercept", multiverse_sign = "positive", multiverse_p = 0.01)` models with a positive one. For depression, of the `r function_dynamic_text_results("multiverse","depression", "intercept", multiverse_sign = "positive", multiverse_p = 0.01) + function_dynamic_text_results("multiverse","depression", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)` models with a statistically significant level shift estimate, all `r  function_dynamic_text_results("multiverse","depression", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)` models returned a negative level shift coefficient.

@@

```{r fig_international, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 7.5, fig.height = 6.5, fig.align='center', out.width = '95%',  fig.scap="CITS effect estimates and confidence intervals by mental health symptom, all model combinations.", out.extra='',  fig.cap="\\label{fig_international}Controlled interrupted time series effect estimates by mental health related symptom for each model specification combination in multiverse analysis. Each point presents the effect estimates of a sinlge model; for each symptom, the estimated slope change per day is shown on the x-axis, while the level shift over the post-announcement period is shown on the y-axis. ADHD: Attention deficit hyperactivity disorder. CoC: clouding of consciousness. GAD: generalized anxiety disorder. MDD: Major depressive disorder."}

@@

* Shows that event the announcement of a safe and effective vaccine led to a significant shift in searches. If searches are taken as a reasonable proxy of mental health status, the period following the announcement saw both a sharp decline in depression and anxiety

* Lack of heterogeneity by polity is interesting, given evidence that polity is a strong predictor of attitudes towards COVID-19 vaccination. Also makes bias from an uncommon shock (e.g., 2020 election) less likely to explain the results

>>>>

    text = element_text(family = "Source Sans Pro")) +
  scale_color_manual(values = c(col_mh,
                                col_headache,
                                col_grey_light),
                     name = "") +
  labs(x = "Slope change", y = "Level shift") +
  geom_text_repel(data=subset(model_outputs_its_graph, estimate_slope < -0.04 |
                                (estimate_level <.1 & mh_indicator ==  "Mental symptom" |
                                   mh_indicator == "Headache (control)")), max.overlaps = 10,
                  size = 3)

@@

While our ITS analysis provided suggestive evidence that search density for mental health symptoms might have changed in the post-announcement period, it also highlighted that searches may have generally changed across other diseases at the same time. Our CITS estimates of each mental health symptom (Figure \ref{figure_cits_outputs}) using searches for headaches as a comparison broadly echoed our ITS analysis, with most conditions showing level and slope changes close to the null. For three conditions, our CITS analysis showed relatively large slope changes with searches for anxiety `r function_dynamic_text_results("main","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("main","depression", "slope", value_include = "allnumber")`, and MDD `r function_dynamic_text_results("main","major_depressive_disorder", "slope", value_include = "allnumber")`, all showing sustained declines. All three of these symptoms had relatively large slope changes in our ITS analysis. In addition, the CITS level shift (i.e., the relative immediate effect) for anxiety was `r function_dynamic_text_results("main","anxiety", "level")`, a decrease double in magnitude compared to the majority of estimates for other symptoms.

@@

```{r figures_cits_data_anxiety, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for anxiety, by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_anxiety}Density of Google Search trends for anxiety, by region and as a function of time. In each row, search density data for two symptoms are displayed: anxiety in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r figures_cits_data_depression, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for depression, by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_depression}Density of Google Search trends for depression, by region and as a function of time. In each row, search density data for two symptoms are displayed: depression in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r figures_cits_data_mdd, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for major depressive disorder (MDD), by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_mdd}Density of Google Search trends for MDD, by region and as a function of time. In each row, search density data for two symptoms are displayed: MDD in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

Slope changes and level shifts for anxiety, depression, and MDD are reported for each combination (144 models) in Figure \ref{fig_multiverse_analysis}. Examining the plausible model combinations of the multiverse analysis for these three symptoms underscored the importance of regional level to inference, as the use of county-level data (with a greater number of observations) consistently showed similar effect estimates to state level estimates for each CITS estimand, but with much tighter confidence intervals, even when geographic fixed effects and narrower bandwidths were included.  Across all models, the sustained change in search density for anxiety and depression is negative and significant an alpha of 0.01. In terms of the estimated level shifts, `r function_dynamic_text_results("multiverse","anxiety", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)` models for anxiety (`r round(function_dynamic_text_results("multiverse","anxiety", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)*100/144,0)`%) had a statistically significant negative change, compared to `r function_dynamic_text_results("multiverse","anxiety", "intercept", multiverse_sign = "positive", multiverse_p = 0.01)` models with a positive one. For depression, of the `r function_dynamic_text_results("multiverse","depression", "intercept", multiverse_sign = "positive", multiverse_p = 0.01) + function_dynamic_text_results("multiverse","depression", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)` models with a statistically significant level shift estimate, all `r  function_dynamic_text_results("multiverse","depression", "intercept", multiverse_sign = "negative", multiverse_p = 0.01)` models returned a negative level shift coefficient.

@@

```{r fig_international, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 7.5, fig.height = 6.5, fig.align='center', out.width = '95%',  fig.scap="International CITS effect estimates and confidence intervals by mental health symptom.", out.extra='',  fig.cap="\\label{fig_international}International controlled interrupted time series effect estimates of changes in search density by mental health related symptom, reported across six nations (Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States). The top panel reports estimates of the level shift and the lower panel reports the estimates for the slope change. Effect estimates are represented as solid black lines, with the shaded regions above and below the lines showing the 95% confidence intervals. The solid black line across all panels reflects a null effect. MDD: Major depressive disorder."}

@@

This study sought to evaluate whether two contemporary announcements of COVID-19 vaccine safety and efficacy on November 9th and November 16th, 2020 were associated with changes in health-related Google search density related to symptoms concerning mental health. Using search data from U.S. counties, states, and international trends, we applied an ITS and CITS modeling approach to investigate associations between the announcements and changes in search density. We found evidence of statistically significant immediate and sustained declines in searches for anxiety, depression, and major depressive disorder throughout the 40 days following the announcements of vaccine safety and efficacy. These declines were relatively large compared to changes in other symptoms, both physical and mental, and a multiverse analysis indicated that they were robust to alternative specifications.

To interpret these observed changes as reactions to the vaccine safety and efficacy announcements, it is important to evaluate whether they may be responses to other contemporary events. We highlight two potential alternatives that coincided with the period under study: the 2020 U.S. general election on November 3rd and the observance of the Thanksgiving holiday in the United States on November 26th. While it is possible that our estimates encompass possible impacts from these events, data from our subgroup analysis lends support for our interpretation that these alternatives may not be sufficient to explain the observed declines in search trends for anxiety, depression, and MDD.

With respect to the U.S. general election, we would anticipate any post-election decline in mental health related symptoms to be concentrated in the United States, given its domestic importance and the political fervor that surrounded it. Similarly, we would expect that impacts of the Thanksgiving holiday would be unique to the U.S., as it is not celebrated by the other nations in the sample and there were no other national holidays observed both those nations at the same time (with the exception of Singapore’s observance of Deepavali on November 14).

In our subgroup analysis comparing search trends across Australia, Ireland, New Zealand, Singapore, the United Kingdom, and the United States, we do not observe evidence of U.S.-specific effect, with our analysis showing broad similarities in estimates of declines in searches related to anxiety and depression. Furthermore, the estimated declines in search density in Australia and the United Kingdom—which, along with the U.S., reflect the more populous nations in the sample—were greater than that of the U.S. for these two symptom searches, providing additional counterevidence to these alternative explanations.

Further countering the explanation that observed declines in search trends for mental health symptoms were due to the U.S. general election are the results from our subgroup analysis comparing blue and red states and counties. If voters expressed relief following the election if their candidate was the winner, a greater decline in searches related to anxiety and depression in states and counties would be expected in those that more heavily voted for Joe Biden, the Democratic candidate and subsequent President-elect, than for former President Donald Trump, the Republic candidate who lost his bid for reelection.  This gradient would also echo our hypothesis that, given findings that political ideology may affect receptiveness to vaccination, the announcement of vaccine safety and efficacy would have had a less pronounced effect among conservative voters compared to liberal ones. Instead, we observed the opposite: among the counties that most heavily voted for the Republican presidential candidate, the estimated declines in search density for anxiety and depression were larger than in those that most heavily voted Democrat. We do, however, observe broad similarities in the confidence intervals of these estimates, such that we do not rule out the possibility that the effects were the same across red and blue counties. However, even this finding would be striking, giving the enduring political controversy over vaccination in the U.S.; while subsequent attitudes towards vaccination may show an ideological split by political affiliation, our findings—if taken as a reasonably reflection of individual attitudes—suggest that whatever relief the vaccine announcements brought was shared across the aisle.

The aggregated nature of our data, however, imposes a constraint to our interpretations, as we are unable to identify whether individual political ideologies were correlated with changes in search trends. Instead, we rely on population aggregates that reflect political orientation (i.e., vote share by county and state) as a proxy, focusing on the 20% most Democrat and Republican-voting areas to strengthen the plausibility of this proxy.

Additionally, we highlight that the use of Google search density as a proxy for population mental health could not be directly validated in our study, as the search data is anonymized and aggregated. While this is an important limitation, prior work has found a strong correlation between searches for anxiety and depression and U.S. population survey responses for receiving therapy, receiving medication for mental health disorders, and having unmet mental health needs. Given the paucity of high frequency population mental health data, Google search density data offered a unique opportunity to investigate the immediate and sustained impacts of vaccine safety and efficacy announcements with an aggregated measures that may be linked with subjective wellbeing and care seeking.

Despite these limitations, our findings suggest that the announcement of COVID-19 vaccine safety and efficacy led to a robust and sustained reduction in Google search density for anxiety and depression, relative to search density prior to the announcements and to searches for other symptoms. If search density echoes population mental health, our results would indicate that these announcements-as a hopeful signal of a new phase of the COVID-19 pandemic-led to an improvement in mental health, underscoring the importance of timely and transparent communication in public health.
